---
title: "Data Project Presentation"
author: "Klerisson Paixao"
date: "Thursday, January 22, 2015"
output: ioslides_presentation
runtime: shiny
mode: selfcontained
url: {lib: ../libraries}
---

## Shiny Presentation

This R Markdown presentation about the interactive NOOA data set.
Goal of this report is to idenify severe weather events of the highest impact on both US population and US economy based on the information retrieved from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

## Data set
Database version used for this report contain entries for years 1950-2011. 
See following the five firts rows:

```{r echo=FALSE,message=FALSE}
library(ggplot2)
library(rCharts)
library(ggvis)

library(data.table)
library(reshape2)
library(dplyr)
library(markdown)
library(mapproj)
library(maps)
dt <- fread('data.csv') %>% mutate(EVTYPE = tolower(EVTYPE))
```

```{r}
head(dt, 5)
```

## NOOA Events plot

```{r, echo=FALSE,message=FALSE}



inputPanel( 
sliderInput("range", 
            "Range:", 
            min = 1950, 
            max = 2011, 
            value = c(1993, 2011),
            sep = ""
        )
)

aggregate_by_state <- function(dt, year_min, year_max, evtypes) {
  replace_na <- function(x) ifelse(is.na(x), 0, x)
  round_2 <- function(x) round(x, 2)
  
  states <- data.table(STATE=sort(unique(dt$STATE)))
  
  aggregated <- dt %>% filter(YEAR >= year_min, YEAR <= year_max, EVTYPE %in% evtypes) %>%
    group_by(STATE) %>%
    summarise_each(funs(sum), COUNT:CROPDMG)
  
  # We want all states to be present even if nothing happened
  left_join(states,  aggregated, by = "STATE") %>%
    mutate_each(funs(replace_na), FATALITIES:CROPDMG) %>%
    mutate_each(funs(round_2), PROPDMG, CROPDMG)    
}

aggregate_by_year <- function(dt, year_min, year_max, evtypes) {
  round_2 <- function(x) round(x, 2)
  
  # Filter
  dt %>% filter(YEAR >= year_min, YEAR <= year_max, EVTYPE %in% evtypes) %>%
    # Group and aggregate
    group_by(YEAR) %>% summarise_each(funs(sum), COUNT:CROPDMG) %>%
    # Round
    mutate_each(funs(round_2), PROPDMG, CROPDMG) %>%
    rename(
      Year = YEAR, Count = COUNT,
      Fatalities = FATALITIES, Injuries = INJURIES,
      Property = PROPDMG, Crops = CROPDMG
    )
}

compute_affected <- function(dt, category) {
  dt %>% mutate(Affected = {
    if(category == 'both') {
      INJURIES + FATALITIES
    } else if(category == 'fatalities') {
      FATALITIES
    } else {
      INJURIES
    }
  })
}

compute_damages <- function(dt, category) {
  dt %>% mutate(Damages = {
    if(category == 'both') {
      PROPDMG + CROPDMG
    } else if(category == 'crops') {
      CROPDMG
    } else {
      PROPDMG
    }
  })
}

plot_impact_by_state <- function (dt, states_map, year_min, year_max, fill, title, low = "#fff5eb", high = "#d94801") {
  title <- sprintf(title, year_min, year_max)
  p <- ggplot(dt, aes(map_id = STATE))
  p <- p + geom_map(aes_string(fill = fill), map = states_map, colour='black')
  p <- p + expand_limits(x = states_map$long, y = states_map$lat)
  p <- p + coord_map() + theme_bw()
  p <- p + labs(x = "Long", y = "Lat", title = title)
  p + scale_fill_gradient(low = low, high = high)
}

plot_impact_by_year <- function(dt, dom, yAxisLabel, desc = FALSE) {
  impactPlot <- nPlot(
    value ~ Year, group = "variable",
    data = melt(dt, id="Year") %>% arrange(Year, if (desc) { desc(variable) } else { variable }),
    type = "stackedAreaChart", dom = dom, width = 650
  )
  impactPlot$chart(margin = list(left = 100))
  impactPlot$yAxis(axisLabel = yAxisLabel, width = 80)
  impactPlot$xAxis(axisLabel = "Year", width = 70)
  
  impactPlot
}

plot_events_by_year <- function(dt, dom = "eventsByYear", yAxisLabel = "Count") {
  eventsByYear <- nPlot(
    Count ~ Year,
    data = dt,
    type = "lineChart", dom = dom, width = 650
  )
  
  eventsByYear$chart(margin = list(left = 100))
  eventsByYear$yAxis( axisLabel = yAxisLabel, width = 80)
  eventsByYear$xAxis( axisLabel = "Year", width = 70)
  eventsByYear
}

states_map <- map_data("state")
evtypes <- sort(unique(dt$EVTYPE))

# Define and initialize reactive values
    values <- reactiveValues()
    values$evtypes <- evtypes
 # Prepare dataset for maps
    dt.agg <- reactive({
        aggregate_by_state(dt, input$range[1], input$range[2], evtypes)
    })

renderPlot({
        print(plot_impact_by_state(
            dt = compute_damages(dt.agg(), 'both'),
            states_map = states_map, 
            year_min = input$range[1],
            year_max = input$range[2],
            title = "Economic impact %d - %d (Million USD)",
            fill = "Damages"
        ))
    })

```

## Results

Most of the observed differences can be attributed to the differences in the size of the samples. Nevertheless I was able to identift four types of events which had both high normalized average and maximum impact and should be considered highly harmful:

- Surge
- Heat / Heat Wave
- Tropical storm
- Tsunami
